---
title: "Monarchs"
author: "Petula Law"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `monarchs.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator:Comma  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=FALSE}
# FILL IN THE CODE BELOW and review the outputs
monarchs1 <- read.csv("data/Monarchs.csv")

library(tidyverse)
monarchs2 <- read_csv("data/Monarchs.csv")

monarchs3 <- read.csv2("data/Monarchs.csv")

monarchs4 <- read_csv2( "data/Monarchs.csv")
```
Answer: 

1. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object

      read_csv() and read_csv2()

2. What is the result of running `class()` on the `monarchs` object created with a tidyverse function.

      [1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?

      There are 12 columns

4. Show the dataset so that we can see how R interprets each column
      
```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

library(tidyverse)
monarchs <- read_csv("data/Monarchs.csv")

class(monarchs)

glimpse(monarchs)

```


## Calculate the duration of reign for all the monarchs in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval = TRUE}

monarchs <- monarchs |>
  mutate(
    monarch_name = str_trim(monarch_name),
    start_of_first_reign_year = as.numeric(str_trim(start_of_first_reign_year)),
    end_of_first_reign_year = as.numeric(end_of_first_reign_year)
  )
glimpse(monarchs)

monarchs <- monarchs |>
  mutate(duration = end_of_first_reign_year - start_of_first_reign_year)

monarchs |>
  select(monarch_name, start_of_first_reign_year, end_of_first_reign_year, duration)


# YOUR CODE


```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r eval = TRUE}
monarchs |>
  select(duration) |>
  summarise(average_duration = mean(duration, na.rm = TRUE))

# YOUR CODE
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `monarchs` dataset. Display the result and also  count the resulting rows with `count()`

```{r eval = TRUE}
average_duration <- mean(monarchs$duration, na.rm = TRUE)
longer_than_average <- monarchs |>
  filter(!is.na(duration), duration > average_duration) 

longer_than_average |>
  select(monarch_name, duration) 

longer_than_average |>
  count()


# YOUR CODE
```


## How many days did the three longest-ruling monarchs rule? 

* Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r eval = TRUE}
top3 <- monarchs |>
  filter(!is.na(duration)) |>
  arrange(desc(duration)) |>
  slice(1:3) |>
  mutate(Days = as.numeric (
    as.Date(end_of_first_reign) - as.Date(start_of_first_reign)
  ))

top3 |>
  select(monarch_name, start_of_first_reign_year, end_of_first_reign_year, duration, Days)


# YOUR CODE
```


## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r eval = TRUE}
monarchs_plot <- monarchs |>
  filter(!is.na(duration), !is.na(start_of_first_reign_year), !is.na(end_of_first_reign_year)) |>
  mutate(midyear = start_of_first_reign_year + (end_of_first_reign_year - start_of_first_reign_year) /2)

ggplot(monarchs_plot, aes (x=midyear, y = duration)) +
  geom_point(size = 2, color = "darkred") + 
  geom_smooth(color = "steelblue", fill = "lightblue") + 
  labs (
    title = "Duration of Reign of British Monarchs, 1066-Present",
    subtitle = "Each point is one monarch, plotted at the midpoint of their reign",
    x = "Year",
    y = "Duration of reign (years)"
  ) +
  theme_bw(base_size = 14)


# YOUR CODE
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!